---
title: "Data 605 Homework 14 - Taylor Series"
author: "Jean Jimenez"
date: "2023-12-02"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(ggplot2)
library(Deriv)

```

# Assignment

This week, we'll work out some Taylor Series expansions of popular functions:

1.  $f(x)=\frac{1}{1-x}$

2.  $f(x)=e^x$

3.  $f(x)=\ln(1+x)$

4.  $f(x)=x^\frac{1}{2}$

For each function, only consider its valid ranges as indicated in the notes when you are computing the Taylor Series expansion. Please submit your assignment as an R- Markdown document.

**Taylor Series Expansion Formula**:

$$
f(x) = f(a) + f'(a)(x - a) + \frac{f''(a)}{2!}(x - a)^2 + \frac{f'''(a)}{3!}(x - a)^3 + \cdots
$$

For this Homework, I will make a function that will evaluate and calculate the taylor series of each equation:

## #1

$f(x)=\frac{1}{1-x}$

Visualization:

```{r, echo=FALSE}

x1 = seq(-10, 10, by = 0.01)

y1 = 1 / (1 - x1)

data1 = data.frame(x1, y1)

ggplot(data1, aes(x1, y1)) + 
    geom_line(color = "red") + 
    ggtitle("Graph of f(x) = 1/(1-x)") +
    xlab("x") + 
    ylab("f(x)") +
    coord_cartesian(xlim=c(-10,10), ylim = c(-10, 10)) +  # Adjust y-axis limits
    theme_minimal()

```

As you can see from the graph, the function $f(x)=\frac{1}{1-x}$ has a vertical asymptote at x=1. When x=1, f(x) is undefined.

$\lim_{{x \to 1 ^-}} f(x)=\infty$

$\lim_{{x \to 1 ^+}} f(x)=-\infty$

Lets do Taylor expansion of f(x) at x=0. I will set a=0. I will be calculating up to the 5th derivative.

```{r}

# Define the original function and its derivatives
f1 = expression(1 / (1 - x))
d1 = D(f1, "x")
d2 = D(d1, "x")
d3 = D(d2, "x")
d4 = D(d3, "x")
d5 = D(d4, "x")

# Evaluate each derivative at x = 0
f1_0 = eval(f1, list(x = 0))
d1_0 = eval(d1, list(x = 0))
d2_0 = eval(d2, list(x = 0))
d3_0 = eval(d3, list(x = 0))
d4_0 = eval(d4, list(x = 0))
d5_0 = eval(d5, list(x = 0))

#Taylor expansion function
taylor_expansion1 = function(x) {
  f1_0 + d1_0 * x + d2_0 * x^2 / factorial(2) + d3_0 * x^3 / factorial(3) +
    d4_0 * x^4 / factorial(4) + d5_0 * x^5 / factorial(5)
}

# original function
f_original1 = function(x) {
  1 / (1 - x)
}

eval1=0

eval1_orig=eval(f_original1(eval1))

eval1_taylor=eval(taylor_expansion1(eval1))

cat("When centered at X=", eval1, "\n the original function outputs:",eval1_orig,". \n", "the taylor expansion outputs:", eval1_taylor,".\n" )
```

## #2

$f(x)=e^x$

Visualization:

```{r, echo=FALSE}

x2 = seq(-10, 10, by = 0.01)

y2 = exp(x2)

data2 = data.frame(x2, y2)

ggplot(data2, aes(x2, y2)) + 
    geom_line(color = "green") + 
    ggtitle("Graph of f(x) = e^x") +
    xlab("x") + 
    ylab("f(x)") +
    coord_cartesian(xlim=c(-10,10), ylim = c(0, 20)) + 
    theme_minimal()

```

As you can see from the graph, the function $f(x)=e^x$ has a horizontal asymptote at y=0. When f(x)=0, x is undefined.

$\lim_{{x \to - \infty }} f(x)=0$

Lets do Taylor expansion of f(x) at x=0. I will set a=0. I will be calculating up to the 5th derivative.

```{r}

# Define the original function and its derivatives
f2 = expression(exp(x))
d21 = D(f2, "x")
d22 = D(d21, "x")
d23 = D(d22, "x")
d24 = D(d23, "x")
d25 = D(d24, "x")

# Evaluate each derivative at x = 0
f2_0 = eval(f2, list(x = 0))
d21_0 = eval(d21, list(x = 0))
d22_0 = eval(d22, list(x = 0))
d23_0 = eval(d23, list(x = 0))
d24_0 = eval(d24, list(x = 0))
d25_0 = eval(d25, list(x = 0))

#Taylor expansion function
taylor_expansion2 = function(x) {
  f2_0 + d21_0 * x + d22_0 * x^2 / factorial(2) + d23_0 * x^3 / factorial(3) +
    d24_0 * x^4 / factorial(4) + d25_0 * x^5 / factorial(5)
}

# original function
f_original2 = function(x) {
  exp(x)
}

eval2=0

eval2_orig=eval(f_original2(eval2))

eval2_taylor=eval(taylor_expansion2(eval2))

cat("When centered at X=", eval2, "\n the original function outputs:",eval2_orig,". \n", "the taylor expansion outputs:", eval2_taylor,".\n" )
```

## #3

$f(x)=\ln(1+x)$

Visualization:

```{r, echo=FALSE}

x3 = seq(-10, 10, by = 0.01)

y3 = log(1+x3)

data3 = data.frame(x3, y3)

ggplot(data3, aes(x3, y3)) + 
    geom_line(color = "purple") + 
    ggtitle("Graph of f(x) = ln(1+x)") +
    xlab("x") + 
    ylab("f(x)") +
    coord_cartesian(xlim=c(-2,10),ylim = c(0, 5)) + 
    theme_minimal()

```

As you can see from the graph, the function $f(x)=\ln(1+x)$ has a verticle asymptote at x=-1. When x=-1, f(x) is undefined.

$\lim_{{x \to - 1^+ }} f(x)=-\infty$

Lets do Taylor expansion of f(x) at x=0. I will set a=0. I will be calculating up to the 5th derivative.

```{r}

# Define the original function and its derivatives
f3 = expression(log(1+x))
d31 = D(f3, "x")
d32 = D(d31, "x")
d33 = D(d32, "x")
d34 = D(d33, "x")
d35 = D(d34, "x")

# Evaluate each derivative at x = 0
f3_0 = eval(f3, list(x = 0))
d31_0 = eval(d31, list(x = 0))
d32_0 = eval(d32, list(x = 0))
d33_0 = eval(d33, list(x = 0))
d34_0 = eval(d34, list(x = 0))
d35_0 = eval(d35, list(x = 0))

#Taylor expansion function
taylor_expansion3 = function(x) {
  f3_0 + d31_0 * x + d32_0 * x^2 / factorial(2) + d33_0 * x^3 / factorial(3) +
    d34_0 * x^4 / factorial(4) + d35_0 * x^5 / factorial(5)
}

# original function
f_original3 = function(x) {
  log(1-x)
}

eval3=0

eval3_orig=eval(f_original3(eval3))

eval3_taylor=eval(taylor_expansion3(eval3))

cat("When centered at X=", eval3, "\n the original function outputs:",eval3_orig,". \n", "the taylor expansion outputs:", eval3_taylor,".\n" )
```

## #4

$f(x)=x^\frac{1}{2}$

Visualization:

```{r, echo=FALSE}

x4 = seq(0, 20, by = 0.01)

y4 = x4^0.5

data4 = data.frame(x4, y4)

ggplot(data4, aes(x4, y4)) + 
    geom_line(color = "brown") + 
    ggtitle("Graph of f(x) = x^(1/2)") +
    xlab("x") + 
    ylab("f(x)") +
    coord_cartesian(xlim=c(0,20),ylim = c(-10, 10)) + 
    theme_minimal()

```

As you can see from the graph, the function $f(x)=x^\frac{1}{2}$ has a vertical asymptote at x=0. When x is negative, f(x) is undefined. This is because the square root of a negative number is imaginary. Function only exists from $[0,\infty)$.

Lets do Taylor expansion of f(x) at x=2. I will be calculating up to the 5th derivative.

```{r}


f_original4 = function(x) x^0.5


d41 = Deriv(f_original4, "x")
d42 = Deriv(d41, "x")
d43 = Deriv(d42, "x")
d44 = Deriv(d43, "x")
d45 = Deriv(d44, "x")
d46 = Deriv(d45, "x")

# Evaluate each derivative at x = 1
eval_point = 1
d41_0 = d41(eval_point)
d42_0 = d42(eval_point)
d43_0 = d43(eval_point)
d44_0 = d44(eval_point)
d45_0 = d45(eval_point)
d46_0 = d46(eval_point)


taylor_expansion4 = function(x) {
  f_original4(eval_point) + 
  d41_0 * (x - eval_point) +
  d42_0 / factorial(2) * (x - eval_point)^2 +
  d43_0 / factorial(3) * (x - eval_point)^3 +
  d44_0 / factorial(4) * (x - eval_point)^4 +
  d45_0 / factorial(5) * (x - eval_point)^5 +
  d46_0 / factorial(6) * (x - eval_point)^6
}

# Evaluate at x = 2
eval_point2 = 2
eval_orig4 = f_original4(eval_point2)
eval_taylor4 = taylor_expansion4(eval_point2)

# Output results
cat("When centered at X =", eval_point2, "\nthe original function outputs:", eval_orig4, ".\n", "The Taylor expansion outputs:", eval_taylor4, ".\n")

```

\
\
